New cardinality estimation algorithms for HyperLogLog sketches

### 2016-06-15 paper/paper.tex
minor changes

Under the Poisson model the log-likelihood and its derivative are given by
\begin{equation}
\label{equ:log_likelihood_single}
\log \mathcal{\symLikelihood}(\symPoissonRate\vert\vec{\symRegVal}) = 
-\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^{\symRegRange}\frac{\symCount_\symRegVal}{2^\symRegVal}+ 
\sum_{\symRegVal=1}^{\symRegRange}\symCount_\symRegVal \log\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)
+
\symCount_{\symRegRange+1} \log\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegRange}}}\right)
\end{equation}
and
\begin{equation}
\frac{d}{d\symPoissonRate}\log \mathcal{\symLikelihood}(\symPoissonRate\vert\vec{\symRegVal}) 
=
-\frac{1}{\symPoissonRate}\left(
\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}}
+\symCount_{\symRegRange+1}\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}}
\right).
\end{equation}
As a consequence, the maximum-likelihood estimate for the Poisson parameter is given by 
\begin{equation}
\symPoissonRateEstimate = \symNumReg\symXEstimate,
\end{equation}
if $\symXEstimate$ is the root of the function
\begin{equation}
\label{equ:funcdef}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\frac{\frac{\symX}{2^\symRegVal}}{1-e^{\frac{\symX}{2^\symRegVal}}}
+\symCount_{\symRegRange+1}\frac{\frac{\symX}{2^\symRegRange}}{1-e^{\frac{\symX}{2^\symRegRange}}}.
\end{equation}
This function can also be written as
\begin{equation}
\label{equ:func}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\left(\frac{\symX}{2^\symRegVal}\right)
+
\symCount_{\symRegRange+1}\symHelper\left(\frac{\symX}{2^\symRegRange}\right)
-
\left(\symNumReg-\symCount_0\right),
\end{equation}
where the function $\symHelper(\symX)$ is defined as
\begin{equation}
\label{equ:helper}
\symHelper(\symX):=1-\frac{\symX}{e^{\symX}-1}.
\end{equation}
$\symHelper(\symX)$ is strictly increasing and concave as can be seen in \cref{fig:helper_function}. For non-negative values $\symX$ the function ranges from $\symHelper(0)=0$ to $\symHelper(\symX\rightarrow \infty)=1$.
\begin{figure}
\centering
\includesvg[width=0.45\textwidth]{helper}
\caption{The function $\symHelper(\symX)$.}
\label{fig:helper_function}
\end{figure}
Since the function $\symFunc(x)$ is also strictly increasing, it is obvious that there exists a unique root $\symXEstimate$ for which $\symFunc(\symXEstimate)=0$. The function is non-positive at 0 since $\symFunc(0)=\symCount_0-\symNumReg\leq 0$ and, in case $\symCount_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing. $\symCount_{\symRegRange+1}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $(\symRegRange+1)$, for which the maximum likelihood estimate would be infinity.

@@

\subsection{Inequalities for the maximum likelihood estimate}
In the following lower and upper bounds for $\symXEstimate$ are derived.
Applying Jensen's inequality on $\symHelper$ in \eqref{equ:func} gives an upper bound for $\symFunc(\symX)$:
\begin{equation}
\symFunc(\symX)
\leq
\symX
\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
+
\left(\symNumReg-\symCount_0\right)\cdot
\symHelper\!\left(
\symX\cdot
\left(
\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
\right)\right)
-
\left(\symNumReg-\symCount_0\right).
\end{equation}
The left-hand side is zero, if $\symXEstimate$ is inserted. Resolution for $\symXEstimate$ finally gives the lower bound
\begin{equation}
\label{equ:strong_lower_bound}
\symXEstimate\geq \frac{\symNumReg-\symCount_0}{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
}
\log\!\left(
1
+
\frac{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
}
{
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
}
\right).
\end{equation}
This bound can be weakened using $\log(1+x) \geq \frac{2x}{x+2}$ for $x\geq0$ which results in
\begin{equation}
\label{equ:weak_lower_bound}
\symXEstimate
\geq
\frac{\symNumReg-\symCount_0}
{\symCount_0+\frac{3}{2}\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCount_\symRegVal}{2^\symRegVal} + \frac{\symCount_{\symRegRange+1}}{2^\symRegRange}}.
\end{equation}
Using the monotonicity of $\symHelper$, the lower bound
\begin{equation}
\symFunc(\symX)
\geq
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)
+
\symCount_{\symRegRange+1}\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)
-
\left(\symNumReg-\symCount_0\right),
\end{equation}
for $\symFunc$ can be found, where $\symRegVal'_\text{max} := \min(\symRegVal_\text{max}, \symRegRange)$ and
$\symRegVal_\text{max} := \max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$.
Again, inserting $\symXEstimate$ and transformation gives
\begin{equation}
\label{equ:strong_upper_bound}
\symXEstimate
\leq
2^{\symRegVal'_\text{max}}
\log\left(
1+
\frac{\symNumReg-\symCount_0}
{
2^{\symRegVal'_\text{max}}
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
}
\right)
\end{equation}
as upper bound which can be weakened using $\log(1+\symX)\leq \symX$ for $\symX\geq 0$
\begin{equation}
\label{equ:weak_upper_bound}
\symXEstimate
\leq
\frac{\symNumReg-\symCount_0}
{\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}}.
\end{equation}
Note, if the HyperLogLog sketch is in the intermediate range, where $\symCount_0=\symCount_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the raw estimate \eqref{equ:intermediate_estimate}. Hence, the maximum likelihood method leads directly to the harmonic mean that is used by the raw estimator. If $\symCount_0>0$ or $\symCount_{\symRegRange+1}>0$ the bounds do no longer follow the harmonic mean and this is also the reason why the raw estimator fails for small and large cardinalities.

@@

\subsection{Maximum-likelihood estimation algorithm}
In order to get a fast cardinality estimation algorithm, it is crucial to minimize the evaluation costs for \eqref{equ:funcprime}. A couple of optimizations allow significant reduction of the computational effort:
\begin{itemize}
\item Only a fraction of all count values $\symCount_\symRegVal$ is non-zero. If we denote $\symRegVal_\text{min}:=\min\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$ and $\symRegVal_\text{max}:=\max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$,  it is sufficient to loop over all indices in the range $[\symRegVal_\text{min}, \symRegVal_\text{max}]$.
\item The coefficient of the linear term $\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}$ can be precalculated and reused for all function evaluations.
\item Many programming languages allow the efficient multiplication and division by any integral power of 2 using special functions, such as \texttt{ldexp} in C/C++ or \texttt{scalb} in Java.
\item The function $\symHelper(\symX)$ only needs to be evaluated at values $\lbrace\frac{\symX}{2^{\symRegVal'_\text{max}}},\frac{\symX}{2^{\symRegVal'_\text{max}-1}}\ldots,\frac{\symX}{2^{\symRegVal_\text{min}}}\rbrace$ where $\symRegVal'_\text{max} := \min(\symRegVal_\text{max}, \symRegRange)$. This series corresponds to a geometric series with ratio 2. A straightforward calculation using \eqref{equ:helper} is very expensive because of the exponential function. However, if we already know $\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
\begin{equation}
\label{equ:helper_recursion1}
\symHelper\left(2\symX\right) = \frac{\symX+2\symHelper\left(\symX\right)\left(1-\symHelper\left(\symX\right)\right)}{\symX+2\left(1-\symHelper\left(\symX\right)\right)}
\end{equation}
or
\begin{equation}
\label{equ:helper_recursion2}
\symHelper\left(\frac{\symX}{2^{\symRegVal}}\right) = \frac{\frac{\symX}{2^{\symRegVal+2}}+\symHelper\left(\frac{\symX}{2^{\symRegVal+1}}\right)\left(1-\symHelper\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}{\frac{\symX}{2^{\symRegVal+2}}+\left(1-\symHelper\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}
\end{equation}
Note, this recursive formula is stable in a sense that the relative error of $\symHelper\left(2\symX\right)$ is smaller than that of $\symHelper\left(\symX\right)$ as shown in \cref{app:helper_stable}.

\item If $\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is smaller than 0.5, the function $\symHelper(\symX)$ can be well approximated by a Taylor series around $\symX=0$
\begin{equation}
\symHelper(\symX)
=
\frac{\symX}{2} - \frac{\symX^2}{12} +\frac{\symX^4}{720}-\frac{\symX^6}{30240} + \symBigO(\symX^{8})
\end{equation}
which can be optimized for numerical evaluation using Estrin's scheme and $\symX' := \frac{\symX}{2}$ and $\symX'' := \symX' \symX'$
\begin{equation}
\label{equ:taylor}
\symHelper(\symX)
=
\symX' - \symX''/3 + \left(\symX'' \symX''\right)\left(1/45-\symX''/472.5\right)
+ \symBigO(\symX^{8})
\end{equation}
In fact, $\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)\leq 0.5$ is almost always fulfilled as long as registers are not saturated. Using \eqref{equ:strong_upper_bound} it is straightforward to see that $\frac{\symX}{2^{\symRegVal'_\text{max}}} \leq \log 2 \approx 0.693$, if $\symCount_{\symRegRange+1}=0$.
In case $\frac{\symX}{2^{\symRegVal'_\text{max}}} > 0.5$, the value $\frac{\symX}{2^{\symE+1}}$ is taken instead, where $\symE$ is the exponent of the floating point representation of $\symX$, $\symE = 1+\lfloor\log_2(\symX)\rfloor$. By definition, $\frac{\symX}{2^{\symE+1}}\leq 0.5$ which allows using the Taylor approximation. $\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is finally obtained after $\symE + 1 -\symRegVal'_\text{max}$ iterations using \eqref{equ:helper_recursion2}.
\end{itemize}

@@

\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\left(\frac{\symXEstimate}{2^\symRegVal}\right)

@@

\symCount_{\symRegRange+1}\symHelper\left(\frac{\symXEstimate}{2^\symRegRange}\right)

@@

\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelperApprox\left(\frac{\symXEstimate\left(1+\symError_\symX\right)}{2^\symRegVal}\right)

@@

\symHelperApprox\left(

@@

\symHelper\left(\frac{\symXEstimate\left(1+\symError_\symX\right)

@@

\symHelper\left(

@@

\symHelper\left(\frac{\symXEstimate

>>>>

Under the Poisson model the log-likelihood and its derivative are given by
\begin{equation}
\label{equ:log_likelihood_single}
\log \mathcal{\symLikelihood}(\symPoissonRate\vert\vec{\symRegVal}) = 
-\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^{\symRegRange}\frac{\symCount_\symRegVal}{2^\symRegVal}+ 
\sum_{\symRegVal=1}^{\symRegRange}\symCount_\symRegVal \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)
+
\symCount_{\symRegRange+1} \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegRange}}}\right)
\end{equation}
and
\begin{equation}
\frac{d}{d\symPoissonRate}\log \mathcal{\symLikelihood}(\symPoissonRate\vert\vec{\symRegVal}) 
=
-\frac{1}{\symPoissonRate}\left(
\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}}
+\symCount_{\symRegRange+1}\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}}
\right).
\end{equation}
As a consequence, the maximum-likelihood estimate for the Poisson parameter is given by 
\begin{equation}
\symPoissonRateEstimate = \symNumReg\symXEstimate,
\end{equation}
if $\symXEstimate$ is the root of the function
\begin{equation}
\label{equ:funcdef}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\frac{\frac{\symX}{2^\symRegVal}}{1-e^{\frac{\symX}{2^\symRegVal}}}
+\symCount_{\symRegRange+1}\frac{\frac{\symX}{2^\symRegRange}}{1-e^{\frac{\symX}{2^\symRegRange}}}.
\end{equation}
This function can also be written as
\begin{equation}
\label{equ:func}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\!\left(\frac{\symX}{2^\symRegVal}\right)
+
\symCount_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^\symRegRange}\right)
-
\left(\symNumReg-\symCount_0\right),
\end{equation}
where the function $\symHelper(\symX)$ is defined as
\begin{equation}
\label{equ:helper}
\symHelper(\symX):=1-\frac{\symX}{e^{\symX}-1}.
\end{equation}
$\symHelper(\symX)$ is strictly increasing and concave as can be seen in \cref{fig:helper_function}. For non-negative values $\symX$ the function ranges from $\symHelper(0)=0$ to $\symHelper(\symX\rightarrow \infty)=1$.
\begin{figure}
\centering
\includesvg[width=0.45\textwidth]{helper}
\caption{The function $\symHelper(\symX)$.}
\label{fig:helper_function}
\end{figure}
Since the function $\symFunc(x)$ is also strictly increasing, it is obvious that there exists a unique root $\symXEstimate$ for which $\symFunc(\symXEstimate)=0$. The function is non-positive at 0 since $\symFunc(0)=\symCount_0-\symNumReg\leq 0$ and, in case $\symCount_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing. $\symCount_{\symRegRange+1}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $(\symRegRange+1)$, for which the maximum likelihood estimate would be infinity.

@@

\subsection{Inequalities for the maximum likelihood estimate}
In the following lower and upper bounds for $\symXEstimate$ are derived.
Applying Jensen's inequality on $\symHelper$ in \eqref{equ:func} gives an upper bound for $\symFunc(\symX)$:
\begin{equation}
\symFunc(\symX)
\leq
\symX
\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
+
\left(\symNumReg-\symCount_0\right)\cdot
\symHelper\!\left(
\symX\cdot
\left(
\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
\right)\right)
-
\left(\symNumReg-\symCount_0\right).
\end{equation}
The left-hand side is zero, if $\symXEstimate$ is inserted. Resolution for $\symXEstimate$ finally gives the lower bound
\begin{equation}
\label{equ:strong_lower_bound}
\symXEstimate\geq \frac{\symNumReg-\symCount_0}{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
}
\log\!\left(
1
+
\frac{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
}
{
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
}
\right).
\end{equation}
This bound can be weakened using $\log(1+x) \geq \frac{2x}{x+2}$ for $x\geq0$ which results in
\begin{equation}
\label{equ:weak_lower_bound}
\symXEstimate
\geq
\frac{\symNumReg-\symCount_0}
{\symCount_0+\frac{3}{2}\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCount_\symRegVal}{2^\symRegVal} + \frac{\symCount_{\symRegRange+1}}{2^\symRegRange}}.
\end{equation}
Using the monotonicity of $\symHelper$, the lower bound
\begin{equation}
\symFunc(\symX)
\geq
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)
+
\symCount_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)
-
\left(\symNumReg-\symCount_0\right),
\end{equation}
for $\symFunc$ can be found, where $\symRegVal'_\text{max} := \min(\symRegVal_\text{max}, \symRegRange)$ and
$\symRegVal_\text{max} := \max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$.
Again, inserting $\symXEstimate$ and transformation gives
\begin{equation}
\label{equ:strong_upper_bound}
\symXEstimate
\leq
2^{\symRegVal'_\text{max}}
\log\left(
1+
\frac{\symNumReg-\symCount_0}
{
2^{\symRegVal'_\text{max}}
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
}
\right)
\end{equation}
as upper bound which can be weakened using $\log(1+\symX)\leq \symX$ for $\symX\geq 0$
\begin{equation}
\label{equ:weak_upper_bound}
\symXEstimate
\leq
\frac{\symNumReg-\symCount_0}
{\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}}.
\end{equation}
Note, if the HyperLogLog sketch is in the intermediate range, where $\symCount_0=\symCount_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the raw estimate \eqref{equ:intermediate_estimate}. Hence, the maximum likelihood method leads directly to the harmonic mean that is used by the raw estimator. If $\symCount_0>0$ or $\symCount_{\symRegRange+1}>0$ the bounds do no longer follow the harmonic mean and this is also the reason why the raw estimator fails for small and large cardinalities.

@@

\subsection{Maximum-likelihood estimation algorithm}
In order to get a fast cardinality estimation algorithm, it is crucial to minimize the evaluation costs for \eqref{equ:funcprime}. A couple of optimizations allow significant reduction of the computational effort:
\begin{itemize}
\item Only a fraction of all count values $\symCount_\symRegVal$ is non-zero. If we denote $\symRegVal_\text{min}:=\min\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$ and $\symRegVal_\text{max}:=\max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$,  it is sufficient to loop over all indices in the range $[\symRegVal_\text{min}, \symRegVal_\text{max}]$.
\item The coefficient of the linear term $\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}$ can be precalculated and reused for all function evaluations.
\item Many programming languages allow the efficient multiplication and division by any integral power of 2 using special functions, such as \texttt{ldexp} in C/C++ or \texttt{scalb} in Java.
\item The function $\symHelper(\symX)$ only needs to be evaluated at values $\left\lbrace\frac{\symX}{2^{\symRegVal'_\text{max}}},\frac{\symX}{2^{\symRegVal'_\text{max}-1}}\ldots,\frac{\symX}{2^{\symRegVal_\text{min}}}\right\rbrace$ where $\symRegVal'_\text{max} := \min(\symRegVal_\text{max}, \symRegRange)$. This series corresponds to a geometric series with ratio 2. A straightforward calculation using \eqref{equ:helper} is very expensive because of the exponential function. However, if we already know $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
\begin{equation}
\label{equ:helper_recursion1}
\symHelper\!\left(2\symX\right) = \frac{\symX+2\symHelper\!\left(\symX\right)\left(1-\symHelper\!\left(\symX\right)\right)}{\symX+2\left(1-\symHelper\!\left(\symX\right)\right)}
\end{equation}
or
\begin{equation}
\label{equ:helper_recursion2}
\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right) = \frac{\frac{\symX}{2^{\symRegVal+2}}+\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}{\frac{\symX}{2^{\symRegVal+2}}+\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}
\end{equation}
Note, this recursive formula is stable in a sense that the relative error of $\symHelper\!\left(2\symX\right)$ is smaller than that of $\symHelper\!\left(\symX\right)$ as shown in \cref{app:helper_stable}.

\item If $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is smaller than 0.5, the function $\symHelper(\symX)$ can be well approximated by a Taylor series around $\symX=0$
\begin{equation}
\symHelper(\symX)
=
\frac{\symX}{2} - \frac{\symX^2}{12} +\frac{\symX^4}{720}-\frac{\symX^6}{30240} + \symBigO(\symX^{8})
\end{equation}
which can be optimized for numerical evaluation using Estrin's scheme and $\symX' := \frac{\symX}{2}$ and $\symX'' := \symX' \symX'$
\begin{equation}
\label{equ:taylor}
\symHelper(\symX)
=
\symX' - \symX''/3 + \left(\symX'' \symX''\right)\left(1/45-\symX''/472.5\right)
+ \symBigO(\symX^{8})
\end{equation}
In fact, $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)\leq 0.5$ is almost always fulfilled as long as registers are not saturated. Using \eqref{equ:strong_upper_bound} it is straightforward to see that $\frac{\symX}{2^{\symRegVal'_\text{max}}} \leq \log 2 \approx 0.693$, if $\symCount_{\symRegRange+1}=0$.
In case $\frac{\symX}{2^{\symRegVal'_\text{max}}} > 0.5$, the value $\frac{\symX}{2^{\symE+1}}$ is taken instead, where $\symE$ is the exponent of the floating point representation of $\symX$, $\symE = 1+\lfloor\log_2(\symX)\rfloor$. By definition, $\frac{\symX}{2^{\symE+1}}\leq 0.5$ which allows using the Taylor approximation. $\symHelper\!\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ is finally obtained after $\symE + 1 -\symRegVal'_\text{max}$ iterations using \eqref{equ:helper_recursion2}.
\end{itemize}

@@

\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\!\left(\frac{\symXEstimate}{2^\symRegVal}\right)

@@

\symCount_{\symRegRange+1}\symHelper\!\left(\frac{\symXEstimate}{2^\symRegRange}\right)

@@

\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelperApprox\!\left(\frac{\symXEstimate\left(1+\symError_\symX\right)}{2^\symRegVal}\right)

@@

\symHelperApprox\!\left(

@@

\symHelper\!\left(\frac{\symXEstimate\left(1+\symError_\symX\right)

@@

\symHelper\!\left(

@@

\symHelper\!\left(\frac{\symXEstimate
